Fine-scale Twitter Data Reveal Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-06-08 doc/temp-sentiment.tex

%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures}

@@

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effect is much stronger in poor and black neighborhoods.
\end{abstract}
\begin{document}

@@

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality or no air-conditioning \cite{Samuelson2020Jun}. Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, with these heterogeneities in vulnerability being more pronounced within cities rather than between cities. Indeed, other studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}.

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely indicator of mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of Tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive Tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}. Finally, the mood expressed in Tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. We therefore draw on a data set of 243 million geolocated tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. Controlling for a variety of fixed effects across all Tweet, we find higher temperatures are associated with lower sentiment (i.e., worsening mood; See Fig. \ref{fig:wbgt}).  Sentiment is highest at 5\textdegree C WBGT (a dry bulb temperature typically around 12\textdegree C/54\textdegree F) with the largest declines in sentiment between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Sentiment also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. We ran models using multiple measures of sentiment and find similar results and patterns across all sentiment measures (see Supplement).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
  \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and sentiment from text in Tweets.  As temperatures increase above 5\textdegree C WBGT, sentiment rapidly declines.}
  \label{fig:wbgt}
\end{figure}

@@

We explored the relationship between neighborhood racial group characteristics and temperature and sentiment and find the effects of heat are felt disproportionately in neighborhoods that are majority black (see Fig \ref{fig:hetero}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority black neighborhoods decrease four times as much as the sentiment of people in other neighborhoods. Additionally, at modest to moderate temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

Exploring the combined effects of neighborhood percent minority race and income in moderating the impacts of heat on mental health, we find separate effects for these two variables (see Supplement). In other words, neither neither percent minority population nor income alone account for the heterogeneity in vulnerability as neighborhoods that are both poorer and higher percent black are more affected by heat than than neighborhoods that are just poorer or higher proportion black.

@@

\subsection*{Sentiment Comparison With Other Events}
Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While a variety of related metrics can be used to quantify sentiment, we use the VADER metric that was specifically designed for microblogs like Twitter \cite{hutto2014vader}.  To give context to this novel metric, we compare the impacts of heat waves, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), to two other changes that have large impacts on sentiment (See Fig. \ref{fig:compare}).  First, we examine the change in average sentiment from Saturday to Monday, the highest and lowest days for Twitter sentiment as well as other mental health metrics, such as suicide rates \cite{CDC2021}.  We also examine the decrease in sentiment associated with one of the most expensive hurricanes in the last decade in the United States: Hurricane Sandy.  Specifically, we compare the mean sentiment of counties affected by Sandy on the week after landfall to the mean sentiment for the week previous.  Hurricanes like Sandy are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.66\linewidth]{../res/comparison_plot.png}
  \caption{Decline in sentiment during a heat wave across several different neighborhoods, compared with other impacts on sentiment, such as major hurricanes, as well as the weekly variation in sentiment from the peak on Saturday to the low on Monday.}
  \label{fig:compare}
\end{figure}

We find that the impacts of heatwaves on sentiment in rich and majority white neighborhoods are less than the average weekly change in sentiment from Saturday to Monday, but the effects of heat waves on majority black or poor neighborhoods are much greater than the average weekly change in sentiment.  Additionally, the impacts of heat waves on more vulnerable neighborhoods are close in severity to the impact of a major hurricane.

\subsection*{Effects by Combined Statistical Area}

To further explore our findings, we examine the impact of heat on sentiment and sentiment inequalities by Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) with over 1 million people (hereafter: cities).  We examine the impacts of heat alone, as well as the change in size of the gap in sentiment scores between both poor and rich neighborhoods and black and non-black neighborhoods.  For our metric of inequality, we show the size of the gap in sentiment as temperatures increase from optimum temperatures of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 30\textdegree C WBGT (42\textdegree C/108\textdegree F).

Examining the effects of heat on sentiment by city, we find that heat is associated with worsened sentiment in 83.6\% of cities, and the effect is significant in 35.3\% of those.  In no cities does heat significantly increase sentiment, and heat affects sentiment in cities across the US, from Florida to New England to California.

We find that heat increases inequalities in sentiment across income groups for 68.9\% of cities, with 35.7\% of these having a statistically significant effect.  Cities with a significantly unequal effect were found in the south, Midwest, northeast, northwest, and southwest, although they were most common in the mid-Atlantic region.  Additionally, many cities in the southwest had an inverted effect, where higher temperatures actually narrowed the gap in sentiment between rich and poor neighborhoods, including Denver and Oklahoma City, which had a statistically significant inverted effect.

Because we found a much stronger effect of temperature on sentiment in majority black neighborhoods compared to other ethnic groups, we also examined increases in sentiment gaps between neighborhoods in the 95th percentile for black population (77.2\% black) and neighborhoods in the 5th percentile for black population (0\% black).  We excluded cities with a black population less than 5\% of the total population.  We found increasing inequalities in sentiment for more black neighborhoods in 64.7\% of cities, with a significant effect in 21.2\% of those (7 cities).  Additionally, three sunbelt cities had a significant inverted effect.  The patterns of inequalities in the impacts of heat on sentiment by race were similar to inequalities by income: cities with large and significant inequalities were found the mid-Atlantic and Midwest.

\begin{figure}[H]
\centering
  \includegraphics[width=0.5\linewidth]{../res/map_combined.png}
  \label{fig:map}
    \caption{Figure (a) shows the change in sentiment for a 25-degree increase in WBGT for cities with a population of over 1 million.  Figures (b) and (c) show inequalities in impacts of temperature on sentiment for cities with over 1 million people.  The value shown is the predicted change in the size of the gap in VADER score as temperatures increase from 5\textdegree C WBGT to 30\textdegree C WBGT between the 5th and 95 percentile for income (a) and the 95th and 5th percentile for percent black (b).  Cities with a statistically significant effect ($p < 0.05$) are outlined in black.  For racial inequalities (b), only cites where black people make up $> 5\%$ of the population are included.}
\end{figure}

\section*{Discussion}
% Benefits of twitter data/overview of findings
We found that twitter data offers significant advantages for observing environmental effects on human well-being due to its very fine spatio-temporal resolution.  We were able to pair each tweet with the local temperature at the time of the tweet, and found a clear association between heat and worsened sentiment.   Moreover, we were able to show heterogeneities in vulnerability by neighborhood characteristics, overturning previous findings that found no heterogeneities at the county level \cite{Burke2018Aug, Mullins2019Dec}.

% Sentiment is a fuzzy indicator but we have BIG DATA
One limitation of data from twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, a user's expressed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large data set of 243 million tweets, allowing us to control for a wide variety of spatial and temporal effects to estimate population-level changes in sentiment in relation to the weather.

% Compare weekly changes in sentiment to weekly changes in suicides
We found that the change from optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to a heatwave of 25\textdegree C WBGT (36\textdegree C/97\textdegree F) is associated with an overall decrease in the VADER sentiment score of 0.01.  This is similar to the degree of change in sentiment over the course of a week from a Monday nadir (0.1268) to a Saturday peak (0.1372), a weekly change associated with a increase in suicides of 23\% over the course of the study period \cite{CDC2021}.  Thus, while research on linkages between sentiment and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of sentiment and suicides at some time scales.  Moreover, like sentiment, suicides are affected by higher temperatures.  Thus, while twitter sentiment is the only mental health indicator available at fine enough spatial and temporal scales to examine neighborhood effects, these changes in sentiment are representative of real human suffering, and are also very likely indicative of medical outcomes like suicides and hospitalizations, although this analysis cannot speak to those outcomes specifically.

% Discuss the NOVELTY: quote Burke and Mullins, refute their interpretations, people have said XX, but we have shown ~XX.
% Variations in vulnerability also indicates ADAPTATION
Previous work conducted with county-scale data found no heterogeneities in mental health vulnerability by income.  Such analyses have led to conclusions that (1) the mental health effects of climate change will be uniform, least in developed countries, and (2) that adaptation with technologies like air-conditioning will not mitigate the mental health effects of climate change.  For example, Mullins et al. conclude that ``high-temperature events will harm mental health everywhere" and that ``individuals have not been able to successfully reduce the negative effects of higher temperatures on mental health" \cite{Mullins2019Dec}.  Contrary to these analyses, we found large heterogeneities in vulnerability by neighborhood race and income.  This suggests that the mental health effects of climate change will not be uniform and, like other impacts, will fall disproportionately on the poor and vulnerable.  More positively, it does suggest that adaptation is possible, and, for communities with infrastructure and working conditions similar those of high-income Americans, the effects of high temperatures on mental health can be largely mitigated.

% Comment on Race, why no effect on hispanics?
We found that majority black neighborhoods are much more affected by heat than neighborhoods with a majority of any other race.  This is surprising, given that Hispanics can be marginalized in both income and housing.  This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than black Americans, or may also be due to that fact that more marginalized and heat-vulnerable Hispanic people were more likely to tweet in the Spanish language, which we did not include our analysis.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine with respect to heat and mental health.  Finally, while race is correlated with income in the US, we found that race contributed to vulnerability independently of income.  This suggests that racism and patterns of housing discrimination such as a legacy of redlining contribute to making black Americans more vulnerable to heat, event at similar levels of income.

% Interpret Spatial Analysis
In our spatial analysis, we found that heat affects sentiment in cities across the US, suggesting that higher baseline temperatures do not reduce vulnerability.   Additionally, most cities exhibited both racial and income inequalities in the vulnerability of mental health to heat, although the southwest had the weakest inequalities and an occasional inverted effect.  While unusual patterns of heat impacts in the southwest could suggest that heat inequality is related to environmental factors such as humidity, this is unlikely because we used Wet Bulb Globe Temperature, which accounts for how differences in humidity and wind speed affect the human body's ability to cool itself through perspiration.  Thus, this regional differences in vulnerability are likely more due to cultural or infrastructural differences than to different baseline environmental conditions.

@@

% Caveats
While these findings are robust to different model specifications and sentiment metrics, there are some caveats.  For one, we were only able to locate the tweets within the census block where the tweet was sent - we did not to infer where the person sending the tweets typically lived or where they had been.  Many low-income and black people commute during the day to work in the service sector in higher-income areas, so these results may in fact under-estimate the impacts of higher temperatures on mental health for poor and minority individuals.  Moreover, in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are typically found that census block.  For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day.  Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American.  Again, these issues mean we are likely under-estimating the true effect of heat on mental health for poor and black people.  A final issue with using twitter data is that, while twitter is used by more than one in five Americans, twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}.  Nevertheless, we found a large volume of tweets across all neighborhood types.

% NCC doesnt have a "conclusion" section, but this is intended as an overall summary paragraph
While climate change will have widespread and severe impacts on human well-being, it cannot be overemphasized just how highly unevenly these impacts will be distributed.  People with more money, access to aid and infrastructure, and who belong to ethnic groups in power are less affected by climate shocks and natural disasters and more able to adapt \cite{bullard2012wrong}.  While the physical health impacts of these shocks are more visible and easy to measure, the mental health impacts of climate change are also causing severe human suffering and there is no reason to believe they will not also be highly unevenly distributed.  Thus, there are strong theoretical priors behind the hypothesis that low-income and marginalized people are more vulnerable to the mental health impacts of higher temperatures, even though previous work at coarse spatial scales had not found such an effect.  By using fine-scale twitter data, we show that there are indeed stark differences in the mental health impacts of heat among neighborhoods in the United States.  These findings have significant implications for urban planning, climate and environmental justice, and mental health.

\section*{Methods}
\subsection*{Tweets \& Sentiment}
We used data from 243 million geo-located English-language tweets from the continental United States from the years 2009 to 2019.  The Twitter data consists of publicly posted messages, or Tweets, that are short status messages users post to the platform.  We only considered users’ original content, thus did not include retweets in our analysis.  Additionally, we excluded all tweets that contained weather-related terms, to ensure that the sentiment expressed in the tweets was reflective of a user's mental state, and not a commentary on the weather.

We assessed the sentiment expressed in the tweets using the VADER (Valence Aware Dictionary for sentiment Reasoning) sentiment corpus \cite{gilbert_vader_2014}. VADER is a lexicon and rule-based sentiment analysis tool that is specifically attuned to sentiments expressed in social media because of several features, as:

@@

Mean annual income per capita is provided by the ACS, and we standardized this variable so that the values for each year were in 2019 dollars.  For racial categories, we combined the various categories provided by the ACS into four racial groups: non-Hispanic white, non-Hispanic black, Hispanic of any race, and an "other" category for non-Hispanic people who were neither black nor white, such as Asian, Native American, or mixed-race people.

>>>>

%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Twitter Data Reveal Neighborhood Inequalities in Mental Health Effects of High Temperatures}

@@

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effect is much stronger in poor and Black neighborhoods.
\end{abstract}
\begin{document}

@@

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality or no air-conditioning \cite{Samuelson2020Jun}. Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, with these heterogeneities in vulnerability being more pronounced within city regions rather than between city regions. Indeed, other studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}.

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely indicator of mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}. Finally, the mood expressed in tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. We therefore draw on a data set of 243 million geolocated tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. Controlling for a variety of fixed effects across all tweets, we find higher temperatures are associated with lower sentiment (i.e., worsening mood; See Fig. \ref{fig:wbgt}).  Sentiment is highest at 5\textdegree C WBGT (a dry bulb temperature typically around 12\textdegree C/54\textdegree F) with the largest declines in sentiment between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Sentiment also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. We ran models using multiple measures of sentiment and find similar results and patterns across all sentiment measures (see Supplement).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
  \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and sentiment from text in tweets.  As temperatures increase above 5\textdegree C WBGT, sentiment rapidly declines.}
  \label{fig:wbgt}
\end{figure}

@@

We explored the relationship between neighborhood racial group characteristics and temperature and sentiment and find the effects of heat are felt disproportionately in neighborhoods that are majority Black (see Fig \ref{fig:hetero}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority Black neighborhoods decrease four times as much as the sentiment of people in other neighborhoods. Additionally, at modest to moderate temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

Exploring the combined effects of neighborhood percent minority race and income in moderating the impacts of heat on mental health, we find separate effects for these two variables (see Supplement). In other words, neither neither percent minority population nor income alone account for the heterogeneity in vulnerability as neighborhoods that are both poorer and higher percent Black are more affected by heat than than neighborhoods that are just poorer or higher proportion Black.

@@

\subsection*{Sentiment Comparison With Other Events}
We compared the expressed sentiment from exposure to temperatures to other events that have large impacts on sentiment (see Fig. \ref{fig:compare}). In particular, we examine change in sentiment from lower to higher temperatures, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F) and compare to changes in sentiment from Saturday to Monday and Hurricane Sandy the week before and after landfall. Previous work indicates sentiment expressed on Twitter are generally highest on Saturdays and lowest on Mondays, while conversely suicide rates and other metrics of mental health are lowest and highest, respectively \cite{CDC2021}. Hurricane Sandy was one of the most damaging and expensive hurricanes to make landfall in the USA in the last decade and hurricanes are associated with mental health effects including PTSD \cite{Schwartz2017Aug, Schwartz2018May}. Similar disasters have also been associated with subsequent increases in suicides \cite{Krug1998Feb}.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.66\linewidth]{../res/comparison_plot.png}
  \caption{Decline in sentiment during exposure to change in temperature from lower to higher across several different neighborhoods, compared with other impacts on sentiment, such as major hurricanes, as well as the weekly variation in sentiment from the peak on Saturday to the low on Monday.}
  \label{fig:compare}
\end{figure}

We find that the decline in sentiment in wealthier and majority white neighborhoods from exposure to increasing temperatures is less than the average weekly change in sentiment from Saturday to Monday (see Fig. \ref{fig:compare}). Conversely, the effects of exposure to increasing temperatures in poorer and majority Black neighborhoods are much larger than the average weekly change in sentiment. Additionally, the impacts of exposure to these changes in temperature for the more vulnerable neighborhoods are close in severity to the impact on sentiment from experiencing a major hurricane.

\subsection*{Temperature Effects by Combined Statistical Area}
We examine the impact of temperature on sentiment across Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) containing over 1 million people (hereafter: city regions). We find that higher temperatures are associated with declining sentiment in 83.6\% of city regions (see Fig. \ref{fig:map}) and the effect is significant in 35.3\% of these. Furthermore, we find no correlation between higher temperatures and increased sentiment in any of the city regions.

Increasing exposure from lower to higher temperatures - 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 30\textdegree C WBGT (42\textdegree C/108\textdegree F) - increases disparities in sentiment across income groups for 68.9\% of city regions, with 35.7\% of these having a statistically significant effect. City regions with a significantly unequal effect were located across the USA although they were most common in the mid-Atlantic region. Additionally, many city regions in the southwest had an inverted effect, where higher temperatures actually narrowed the gap in sentiment between wealthier and poorer neighborhoods, including Denver and Oklahoma City areas, which had a statistically significant inverted effect.

Given the strong effect of temperature on sentiment in majority Black neighborhoods compared to other non-majority Black neighborhoods, we also examined increases in sentiment gaps between neighborhoods in the 95th percentile for Black population (77.2\% Black) and neighborhoods in the 5th percentile for Black population (0\% Black). We excluded city regions with a Black population less than 5\% of the total population. We found increasing inequalities in sentiment for more Black neighborhoods in 64.7\% of city regions, with a significant effect in 21.2\% of those (7 city regions).  Additionally, three sunbelt city regions had a significant inverted effect. The patterns of inequalities in the impacts of heat on sentiment by race were similar to inequalities by income: city regions with large and significant inequalities were located in the mid-Atlantic and Midwest.

\begin{figure}[H]
\centering
  \includegraphics[width=0.5\linewidth]{../res/map_combined.png}
  \label{fig:map}
    \caption{Figure (a) shows the change in sentiment for a 25-degree increase in WBGT for city regions with a population of over 1 million.  Figures (b) and (c) show inequalities in impacts of temperature on sentiment for city regions with over 1 million people.  The value shown is the predicted change in the size of the gap in VADER score as temperatures increase from 5\textdegree C WBGT to 30\textdegree C WBGT between the 5th and 95 percentile for income (b) and the 95th and 5th percentile for percent Black (c). City regions with a statistically significant effect ($p < 0.05$) are outlined in Black.  For racial inequalities (c), only cities where the proportion of Black people is $> 5\%$ of the population are included.}
\end{figure}

\section*{Discussion}
% Benefits of Twitter data/overview of findings
We find that Twitter data offers significant advantages for observing environmental effects on human well-being due to its very fine spatio-temporal resolution. We were able to pair each tweet with the local temperature at the time of the tweet, and found a clear association between increasing temperatures and declining sentiment. Moreover, we were able to show heterogeneities in vulnerability by neighborhood characteristics, overturning previous findings that found no heterogeneities at the county level \cite{Burke2018Aug, Mullins2019Dec}.

% Compare weekly changes in sentiment to weekly changes in suicides
We found that the change from optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to a higher temperature of 25\textdegree C WBGT (36\textdegree C/97\textdegree F) is associated with an overall decrease in sentiment of 0.01.  This is similar to the degree of change in sentiment over the course of a week from a Monday nadir (0.1268) to a Saturday peak (0.1372), a weekly change associated with an increase in suicides of 23\% over the course of the study period \cite{CDC2021}. Thus, while research on linkages between sentiment and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of sentiment and suicides at some time scales. Moreover, like sentiment, suicides are affected by higher temperatures. Thus, while Twitter sentiment is a mental health indicator at fine enough spatial and temporal scales to examine neighborhood effects, these changes in sentiment are representative of real human suffering, and are also very likely indicative of medical outcomes like suicides and hospitalizations, although our analyses here do not incorporate these outcomes.

% Discuss the NOVELTY: quote Burke and Mullins, refute their interpretations, people have said XX, but we have shown ~XX.
% Variations in vulnerability also indicates ADAPTATION
Previous work conducted with county-scale data found no heterogeneities in mental health vulnerability by income.  Such analyses have led to conclusions that (1) the mental health effects of climate change will be uniform, least in developed countries, and (2) that adaptation with technologies like air-conditioning will not mitigate the mental health effects of climate change.  For example, Mullins et al. conclude that ``high-temperature events will harm mental health everywhere" and that ``individuals have not been able to successfully reduce the negative effects of higher temperatures on mental health" \cite{Mullins2019Dec}. Contrary to these analyses, we found large heterogeneities in vulnerability by neighborhood race and income.  This suggests that the mental health effects of climate change will not be uniform and, like other impacts, will fall disproportionately on the poor and vulnerable.  More positively, it does suggest that adaptation is possible, and, for communities with infrastructure and working conditions similar to those of higher-income Americans, the effects of high temperatures on mental health can be addressed and mitigated.

% Comment on Race, why no effect on hispanics?
We found that majority Black neighborhoods are much more affected by higher temperatures than neighborhoods with a non-Black majority. This is surprising, given that Hispanics can be marginalized in both income and housing. This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than Black Americans, or may also be due to that fact that more marginalized and heat-vulnerable Hispanic people were more likely to tweet in the Spanish language, which we did not include our analysis.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine with respect to heat and mental health.  Finally, while race is correlated with income in the US, we found that race contributed to vulnerability to heat, independently of income. This suggests that racism and patterns of housing discrimination like the legacy of redlining contribute to making Black Americans more vulnerable to heat, event at similar levels of income.

% Interpret Spatial Analysis
In our spatial analysis, we found that higher temperatures affect sentiment in city regions across the USA, suggesting that higher baseline temperatures in warmer regions do not necessarily reduce vulnerability. Additionally, most city regions exhibited both racial and income inequalities in the vulnerability of mental health to high temperatures, although the Southwest had the weakest inequalities and an occasional inverted effect.  While unusual patterns of temperature impacts in the Southwest could suggest that temperature inequality is related to environmental factors such as humidity, this is unlikely because we used Wet Bulb Globe Temperature (WBGT) which accounts for how differences in humidity and wind speed affect the human body's ability to cool itself through perspiration. Thus, this regional differences in vulnerability are likely more due to cultural or infrastructural differences than to different baseline environmental conditions.

@@

% Caveats
% Sentiment is a fuzzy indicator but we have BIG DATA
While these findings are robust to different model specifications and sentiment metrics, there are some caveats and limitations. One limitation of data from Twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, a user's expressed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large data set of 243 million tweets, allowing us to control for a wide variety of spatial and temporal effects to estimate population-level changes in sentiment in relation to the weather.

One caveat associated with this data is that we were only able to locate the tweets within the census block where the tweet was sent - we did not to infer where the person sending the tweets typically lived or where they had been.  Many low-income and Black people commute during the day to work in the service sector in higher-income areas, so these results may in fact under-estimate the impacts of higher temperatures on mental health for poor and minority individuals.  Moreover, in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are typically found that census block.  For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day.  Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American.  Again, these issues mean we are likely under-estimating the true effect of heat on mental health for poor and Black people.  A final issue with using Twitter data is that, while Twitter is used by more than one in five Americans, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}.  Nevertheless, we found a large volume of tweets across all neighborhood types.

% NCC doesnt have a "conclusion" section, but this is intended as an overall summary paragraph
While climate change will have widespread and severe impacts on human well-being, our work emphasizes just how highly unevenly these impacts will be distributed. People with more money, access to aid and infrastructure, and who belong to ethnic groups in power are less affected by climate shocks and natural disasters and more able to adapt \cite{bullard2012wrong}. While the physical health impacts of these shocks are more visible and easy to measure, the mental health impacts of climate change are also causing severe human suffering and there is no reason to believe they will not also be highly unevenly distributed.  Thus, there are strong theoretical priors behind the hypothesis that low-income and marginalized people are more vulnerable to the mental health impacts of higher temperatures, even though previous work at coarse spatial scales had not found such an effect.  By using fine-scale Twitter data, we show that there are indeed stark differences in the mental health impacts of heat among neighborhoods in the United States.  These findings have significant implications for urban planning, climate and environmental justice, and mental health.

\section*{Methods}
\subsection*{Tweets \& Sentiment}
We used data from 243 million geo-located English-language tweets from the continental United States from the years 2009 to 2019.  The Twitter data consists of publicly posted messages, or tweets, that are short status messages users post to the platform.  We only considered users’ original content, thus did not include retweets in our analysis.  Additionally, we excluded all tweets that contained weather-related terms, to ensure that the sentiment expressed in the tweets was reflective of a user's mental state, and not a commentary on the weather.

Expressed sentiment is a widely used metric to assess mental health and well-being from posts on social media platforms.  We use the VADER metric to assess the sentiment of Twitter posts as it was specifically designed for microblogs like Twitter \cite{hutto2014vader}.  The VADER (Valence Aware Dictionary for sentiment Reasoning) sentiment corpus \cite{gilbert_vader_2014} is a lexical, rule-based sentiment analysis tool that is specifically attuned to sentiments expressed on social media and includes several features, as:

@@

Mean annual income per capita is provided by the ACS, and we standardized this variable so that the values for each year were in 2019 dollars.  For racial categories, we combined the various categories provided by the ACS into four racial groups: non-Hispanic white, non-Hispanic Black, Hispanic of any race, and an "other" category for non-Hispanic people who were neither Black nor white, such as Asian, Native American, or mixed-race people.
